Ricci-Ollivier Curvature of two Random Walks on the Rooted Phylogenetic Subtree-Prune-Regraft Graph

### 2015-03-30 curvature.tex
A few more twiddles.

The analysis can be reproduced using the SCons (\url{http://scons.org/}) build tool and an IPython notebook; instructions are in the repository README file.

@@

We now bound the relative and exact difference in rSPR degree between two trees with $n$ leaves.

\begin{lemma}
	\label{lem:degree_max_delta}
	Let $T$ and $S$ be rooted phylogenetic trees with $n \ge 3$ leaves, and assume without loss of generality that $\degree{T} \le \degree{S}$.
	Then:
	\begin{enumerate}
		\item $\frac{\degree{T}}{\degree{S}} \ge \frac{3}{4}$, and
		\item $\degree{S} - \degree{T} \le n^2 - 5n + 6$.
	\end{enumerate}
\end{lemma}
\begin{proof}
	To prove (i), we simply note from Lemma~\ref{lem:degree_extremes} that the ladder tree achieves the minimum degree, and the balanced tree achieves the maximum degree:
	\begin{align*}
		\frac{\degree{T}}{\degree{S}} \ge\ &\frac{3n^2 - 13n + 14}{4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4(n-2)^2 - 2(n-2)} \\
		\cuttable{= \ &\frac{3n^2 - 13n + 12}{4n^2 - 16n + 16 - 2(n-2)} \\}
		= \ &\frac{3n^2 - 13n + 12}{4n^2 - 18n + 20} \\
		\cuttable{\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 17\frac{1}{3}n + 18} &\forall n \ge 3 \\}
	\end{align*}
	which is greater than 3/4 when $n \ge 3$.
    Similarly for (ii):
	\begin{align*}
		\cuttable{\degree{S} - \degree{T}
		\le\ &(4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)})
		- (3n^2 - 13n + 14) \\}
		\le\ &(4(n-2)^2 - 2(n-2)) - (3n^2 - 13n + 14) \\
		\cuttable{=\ &4n^2 - 16n + 16 - 2n +4 - 3n^2 + 13n - 14 \\}
		=\ &n^2 - 5n + 6.
	\end{align*}
\end{proof}

@@

The difference in rSPR degree is thus $2ka - 2kb + 2(i-1) - 2(j-1)$ as claimed.
\end{proof}

@@

\section{Curvature}
For the purposes of this paper, $\kappa_n$ is the Ricci-Ollivier curvature \eqref{eq:curvatureDef} on the $n$-taxon rSPR graph with respect to a specified random walk and $W_{1,n}$ is the corresponding mass transport term.

@@

Now we prove the proposition when the number of neighbors differ.
Assume without loss of generality that $\degree{T_n} < \degree{S_n}$.
By Lemma~\ref{lem:degree_change}, $\degree{S_n} - \degree{T_n} = 2(k(a-b) + i - j)$, where each of $\{a,b,i,j\}$ is less than $n$.
Thus, $\degree{S_n} - \degree{T_n} = \OhOf{kn}$.
We again pair neighbor $T'_n$ of $T$ with neighbor $S'_n$ of $S$ such that $\dspr{T'_n, S'_n}=1$ but, as $\degree{T_n} < \degree{S_n}$ we can only account for at most $\degree{T_n} / \degree{S_n}$ of the mass directly and may have to move the $(\degree{S_n} - \degree{T_n}) / \degree{S_n}$ remainder to trees a distance at most 3.
Thus, $W_{1,n}$ is bounded above by $(\degree{T_n} + \OhOf{kn}) / \degree{S_n} = (\degree{S_n} + \OhOf{kn}) / \degree{S_n}$.
We again bound $W_{1,n}$ from below with $(\degree{T_n} - \OhOf{n}) / \degree{T_n}$ by ignoring the mass in common neighbors of $T_n$ and $S_n$.
The proposition again follows because $\degree{T_n}$ is quadratic in $n$.

Next we note a rough bound on the curvature of two trees with respect to their distance.

@@

We can obtain a tighter bound on the maximum curvature of two adjacent trees.
\begin{lemma}
	The maximum curvature of the uniform random walk between two adjacent trees with $n$ leaves is $\frac{6n-17}{3n^2-13n+14}$.
\end{lemma}
\begin{proof}
The maximum curvature between adjacent trees $T$ and $S$ occurs when their neighborhoods have maximum overlap and all other tree pairs are at distance 1.
By Lemma~\ref{lem:shared_neighbors} the maximum overlap is $6n-17$.
The amount of overlapping mass in the shared neighbors of $T$ and $S$ is thus $\frac{6n-17}{\max(\degree{T},\degree{S})}$.
The minimum mass transfer cost is thus $1 - \frac{6n-17}{\max(\degree{T},\degree{S})}$.
This is minimized when $\degree{T}=\degree{S}$ are as small as possible, that is $T,S$ are ladders and $\degree{T} = 3n^2 - 13n + 14$.

@@

We next bound the effect of laziness on curvature.
The lazy random walk again selects a tree uniformly at random, as with the uniform random walk, but only accepts the move with probability $p$.
We denote the curvature of the lazy random walk between two trees $T$ and $S$ by $\curvature[\ell_p]{T,S}$.
For example, $\curvature[\ell_{\frac{1}{4}}]{T,S}$ describes the curvature of the lazy random walk that moves with probability $\frac{1}{4}$ and remains stationary with probability $\frac{3}{4}$.
For the lazy random walk, $m_T$ is now $T \cup N(T)$, with each neighbor assigned mass $\frac{p}{\degree{T}}$ and $T$ assigned the remaining $1 - p$ mass.
Note that Ollivier defined $\curvature[\ell_p]{T, S} := \left(1 - \frac{W_1(m_T, m_S)}{d(x, y)}\right) / p$~\cite{Ollivier2009-bw}.
This definition of curvature is invariant of $p$ for small enough $p$ and was used by Ollivier to avoid parity problems on graphs where the uniform random walk is periodic.
As we now prove, the curvature of the lazy random walk under this definition is closely related to that of the uniform random walk.

@@

\begin{proof}
	We first prove the lower bound, that is $\curvature{T,S} \le \curvature[\ell_p]{T,S}$.
	Let $W_1(T,S)$ be the mass transport cost in the uniform case, and $W_1'(T,S)$ be the same for the lazy case with parameter $p$.
	Recall that $\curvature{T,S} = \curvature[\ell_1]{T,S} = 1 - \frac{W_1(T,S)}{\dspr{T,S}}$, and $\curvature[\ell_p]{T,S} = (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p$.
	Observe that $$W_1'(T,S) \le pW_1(T,S) + (1-p)\dspr{T,S},$$ by taking an optimal uniform mass assignment spread over $p$ of the mass and moving the remainder from $T$ to $S$ (or vice versa).
	Then:
	\begin{align*}
		\curvature[\ell_p]{T,S} &= (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p \\
		&\ge (1 - \frac{pW_1(T,S) + (1-p)\dspr{T,S}}{\dspr{T,S}}) / p \\
		&\ge (\frac{1}{p} - \frac{W_1(T,S)}{\dspr{T,S}} - \frac{1-p}{p}) \\
		&\ge 1 - \frac{W_1(T,S)}{\dspr{T,S}} \\
		&\ge \curvature{T,S}
	\end{align*}

>>>>

The analysis generating the can be reproduced by simply invoking the SCons (\url{http://scons.org/}) build tool and evaluating cells in an IPython notebook; instructions are in the repository README file.

@@

We now bound the ratio and difference of rSPR degree between two trees with $n$ leaves.
\begin{lemma}
	\label{lem:degree_max_delta}
	Let $T$ and $S$ be rooted phylogenetic trees with $n \ge 3$ leaves, and assume without loss of generality that $\degree{T} \le \degree{S}$.
	Then:
	\begin{enumerate}
		\item $\frac{\degree{T}}{\degree{S}} \ge \frac{3}{4}$, and
		\item $\degree{S} - \degree{T} \le n^2 - 5n + 6$.
	\end{enumerate}
\end{lemma}
\begin{proof}
	To prove (i), we simply note from Lemma~\ref{lem:degree_extremes} that the ladder tree achieves the minimum degree, and the balanced tree achieves the maximum degree:
	\begin{align*}
		\frac{\degree{T}}{\degree{S}} \ge\ &\frac{3n^2 - 13n + 14}{4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4(n-2)^2 - 2(n-2)} \\
		\cuttable{= \ &\frac{3n^2 - 13n + 12}{4n^2 - 16n + 16 - 2(n-2)} \\}
		= \ &\frac{3n^2 - 13n + 12}{4n^2 - 18n + 20} \\
		\cuttable{\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 17\frac{1}{3}n + 18} &\forall n \ge 3 \\}
	\end{align*}
	which is greater than 3/4 when $n \ge 3$.
    Similarly for (ii):
	\begin{align*}
		\cuttable{\degree{S} - \degree{T}
		\le\ &(4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)})
		- (3n^2 - 13n + 14) \\}
		\le\ &(4(n-2)^2 - 2(n-2)) - (3n^2 - 13n + 14) \\
		\cuttable{=\ &4n^2 - 16n + 16 - 2n +4 - 3n^2 + 13n - 14 \\}
		=\ &n^2 - 5n + 6.
	\end{align*}
\end{proof}

@@

Thus the difference in rSPR degree is $2ka - 2kb + 2(i-1) - 2(j-1)$ as claimed.
\end{proof}

@@

\section{Curvature}
For the purposes of this paper, $\kappa_n$ is the Ricci-Ollivier curvature \eqref{eq:curvatureDef} on the $n$-taxon rSPR graph with respect to a specified random walk and $W_{1,n}$ is the corresponding mass transport term from \eqref{eq:curvatureDef}.

@@

Now we prove the proposition when the number of neighbors differ.
Assume without loss of generality that $\degree{T_n} < \degree{S_n}$.
By Lemma~\ref{lem:degree_change}, $\degree{S_n} - \degree{T_n} = 2(k(a-b) + i - j)$, where each of $\{a,b,i,j\}$ is less than $n$.
Thus, $\degree{S_n} - \degree{T_n} = \OhOf{kn}$.
We again pair neighbor $T'_n$ of $T$ with neighbor $S'_n$ of $S$ such that $\dspr{T'_n, S'_n}=1$ but, as $\degree{T_n} < \degree{S_n}$ we can only account for at most $\degree{T_n} / \degree{S_n}$ of the mass directly and may have to move the $(\degree{S_n} - \degree{T_n}) / \degree{S_n}$ remainder to trees a distance at most 3.
Thus, $W_{1,n}$ is bounded above by $(\degree{T_n} + \OhOf{kn}) / \degree{S_n} = (\degree{S_n} + \OhOf{kn}) / \degree{S_n}$.
We again bound $W_{1,n}$ from below with $(\degree{T_n} - \OhOf{n}) / \degree{T_n}$ by ignoring the mass in common neighbors of $T_n$ and $S_n$.
The proposition again follows because $\degree{T_n}$ is quadratic in $n$.
\end{proof}

Next we note a rough bound on the curvature of two trees with respect to their distance, then obtain a tighter bound on the maximum curvature of two adjacent trees.

@@

\begin{lemma}
	The maximum curvature of the uniform random walk between two adjacent trees with $n$ leaves is $\frac{6n-17}{3n^2-13n+14}$.
\end{lemma}
\begin{proof}
The maximum curvature between adjacent trees $T$ and $S$ occurs when their neighborhoods have maximum overlap and all other tree pairs are at distance 1.
By Lemma~\ref{lem:shared_neighbors} the maximum overlap is $6n-17$.
The amount of overlapping mass in the shared neighbors of $T$ and $S$ is thus $\frac{6n-17}{\max(\degree{T},\degree{S})}$.
The minimum mass transfer cost is thus $1 - \frac{6n-17}{\max(\degree{T},\degree{S})}$.
This is minimized when $\degree{T}=\degree{S}$ are as small as possible, that is $T,S$ are ladders and $\degree{T} = 3n^2 - 13n + 14$.

@@

We next bound the effect of laziness on curvature.
The lazy random walk again selects a neighbor uniformly at random, as with the uniform random walk, but only travels to the neighbor with probability $p$.
We denote the curvature of the lazy random walk between two trees $T$ and $S$ by $\curvature[\ell_p]{T,S}$.
For example, $\curvature[\ell_{\frac{1}{4}}]{T,S}$ describes the curvature of the lazy random walk that moves with probability $\frac{1}{4}$ and remains stationary with probability $\frac{3}{4}$.
For the lazy random walk, $m_T$ is now $T \cup N(T)$, with each neighbor assigned mass $\frac{p}{\degree{T}}$ and $T$ assigned the remaining $1 - p$ mass.
Note that Ollivier defined $\curvature[\ell_p]{T, S} := \left(1 - \frac{W_1(m_T, m_S)}{d(x, y)}\right) / p$~\cite{Ollivier2009-bw}.
This definition of curvature is invariant of $p$ for small enough $p$ and was used by Ollivier to avoid parity problems on graphs where the uniform random walk is periodic.
As we now prove, the curvature of the lazy random walk under this definition is closely related to that of the uniform random walk.

@@

\begin{proof}
	We first prove the lower bound, that is $\curvature{T,S} \le \curvature[\ell_p]{T,S}$.
	Let $W_1(T,S)$ be the mass transport cost in the uniform case, and $W_1'(T,S)$ be the same for the lazy case with parameter $p$.
	Recall that $\curvature{T,S} = \curvature[\ell_1]{T,S} = 1 - \frac{W_1(T,S)}{\dspr{T,S}}$, and $\curvature[\ell_p]{T,S} = (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p$.
	Observe that $$W_1'(T,S) \le pW_1(T,S) + (1-p)\dspr{T,S},$$ by the simple mass transport program obtained by treating the mass at $T$ and $S$ as separate from that of the neighbors.
	Then:
	\begin{align*}
		\curvature[\ell_p]{T,S} &= (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p \\
		&\ge (1 - \frac{pW_1(T,S) + (1-p)\dspr{T,S}}{\dspr{T,S}}) / p \\
		&\ge (\frac{1}{p} - \frac{W_1(T,S)}{\dspr{T,S}} - \frac{1-p}{p}) \\
		&\ge 1 - \frac{W_1(T,S)}{\dspr{T,S}} \\
		&\ge \curvature{T,S}
	\end{align*}
